---
title: "Introduction to joinpointR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| id: setup
#| include: false
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Joinpoint regression is commonly used in epidemiology to identify changes in temporal trends. The joinpointR package provides tools to fit joinpoint regression models using the grid search method, estimate annual percentage changes, and present results.

## Joinpoint regression models by group
The function `model_jp_grid()` allows to fit joinpoint regression models with a log-response by levels of up to two categorical variables, using the grid search method and model selection based in the Bayesian Information Criteria (BIC).
For the examples, we are going to use a simulated dataset with the HIV rates by sex in five regions

```{r}
#| message: false
#| id: example-data
# Load required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(joinpointR)

# Load example data
data(hiv_data)
```

### Model selection by one grouping variable

```{r}
#| id: fit-1
## Create a reduced dataset
data_sex <- hiv_data |>
  filter(admin == "ARG")

## Fit the joinpoint model
mods_sex <- model_jp_grid(
  data = data_sex,
  rate = hiv_rate,
  time = year,
  group = "sex"
)
```

### Model selection by two grouping variables
```{r}
#| id: fit-2
## Create a reduced dataset
data_admin <- hiv_data |>
  filter(between(admin, "Buenos Aires", "Chubut"))

## Fit the joinpoint model
mods <- model_jp_grid(
  data = data_admin,
  rate = hiv_rate,
  time = year,
  group = c("admin", "sex")
)
```

## Summarise model results
### Summary tables
The functions `get_summary()` provides a summary table containing:
- `model`: Model name.
- `jp`: The number of fitted joinpoints.
- `period`: The time segments defined by the observed joinpoints. Defaults to the whole period when no joinpoints were detected.
- `apc`,`apc_lower`, `apc_upper`, `apc_sig`: The Annual Percent Change (APC) with its confidence interval and significance stars.
- `aapc`,`aapc_lower`, `aapc_upper`, `apc_sig`: The Average Annual Percent Change (APC) with its confidence interval and significance stars. 

 ```{r}
 #| id: get_summary-1
 get_summary(mods_sex)
 ```

By default, `get_summary()` displays both the APC and AAPC, to show only the APC use:

```{r}
#| id: get_summary-2
get_summary(mods_sex, stats = "apc")

get_apc(mods_sex)
```

To display only the AAPC use:
```{r}
#| id: get_summary-3
get_summary(mods_sex, stats = "aapc")

get_aapc(mods_sex)
```

The confidence interval or significance stars can be changed using the argument `hide`:

```{r}
#|id: get_summary-4
# Hide the confidence interval
get_summary(mods_sex, hide = "ci")

# Hide the significance stars
get_summary(mods_sex, hide = "sig")
```

The confidence level defaults to 95% and can be changed using the argument `level.ci`:
```{r}
#| id: get_summary-5
get_summary(mods_sex, level.ci = .9)
```

Results can also be displayed as a `flextable`:

```{r}
#| id: get_summary-6
get_summary(mods_sex, as.ft = TRUE)
```

