---
title: "Introduction to JoinpointR"
author: "Tamara Ricardo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to JoinpointR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| id: settings
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Joinpoint regression is commonly used in epidemiology to identify changes in temporal trends. The joinpointR package provides tools to fit joinpoint regression models, estimate annual percentage changes, and present results.

## Joinpoint regression models by group
The function `model_jp()`allows to fit joinpoint regression models with a log-response by levels of one or more categorical variable(s).
For the examples, we are going to use a simulated dataset with the HIV rates by sex in five regions

```{r}
#| id: example-data
# Load required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(joinpointR)

# Seed for reproducibility
set.seed(123)

# Generate example data
data <- expand_grid(
  year = 2015:2024,
  sex = c("Male", "Female"),
  region = c("North", "Central", "South", "East", "West")
) %>%

  left_join(
    tribble(
      ~region   , ~slope1 , ~slope2 ,
      "North"   ,      -6 ,       4 ,
      "Central" ,      -5 ,       3 ,
      "South"   ,       3 ,      -4 ,
      "East"    ,      -7 ,       2 ,
      "West"    ,       2 ,      -5
    )
  ) %>%

  mutate(
    trend = if_else(
      year <= 2019,
      slope1 * (year - 2015),
      slope1 * 4 + slope2 * (year - 2019)
    ),

    hiv_rate = 100 +
      trend +
      if_else(sex == "Male", 10, 0) +
      rnorm(n(), 0, 0.8)
  ) %>%

  select(year, region, sex, hiv_rate)
```

### Stepwise joinpoint regression by sex
```{r}
#| id: model-1
mod1 <- model_jp(
  data = data,
  value = "hiv_rate",
  time = "year",
  group = "sex",
  k = 3,
  step = TRUE,
  test = TRUE
)
```

### Fixed number of joinpoints by sex
```{r}
#| id: model-2
mod2 <- model_jp(
  data = data,
  value = "hiv_rate",
  time = "year",
  group = "sex",
  k = 1,
  step = FALSE,
  test = FALSE
)

mod2
```

### Stepwise joinpoint regression models by sex and region
```{r}
#| id: model-3
mod3 <- model_jp(
  data = data,
  value = "hiv_rate",
  time = "year",
  group = c("sex", "region"),
  step = TRUE
)
```

## Annual percentage change (APC)
The function `get_apc()` provides the annual percentage change and its 95% confidence interval for models with significant joinpoints (`class: "segmented" "lm"`)

```{r}
#| id: APC-1
get_apc(mod2$Female)
```

The function returns an error when no significant joinpoints are detected (`class = "lm"`).

```{r}
#| error: true
#| id: APC-2
get_apc(mod1$Female, digits = 1)
```

## Average annual percent change (AAPC)
The function `get_aapc()` allows users to estimate the global trend even when no significant joinpoints were detected.

```{r}
#| id: AAPC-1
get_aapc(mod1$Female, digits = 1, show_ci = TRUE)
```

You can choose to show significance stars instead of CI:
```{r}
#| id: AAPC-2
get_aapc(mod1$Female, digits = 1, show_ci = FALSE)
```

## Summary tables
### Results as a data frame
```{r}
# id: tab-1
mod2 |>
  summary_jp(digits = 1, var1 = "Sex", ft = FALSE)
```

### HTML Tables
```{r}
# id: tab-2
mod2 |>
  summary_jp(digits = 1, var1 = "Sex", ft = TRUE)
```

### HTML Tables by two grouping variables
```{r}
#| id: tab-3
mod3 |>
  summary_jp(digits = 1, var1 = "Region", var2 = "Sex", ft = TRUE)
```

Change the table language to Spanish
```{r}
#| id: tab-4
mod3 |>
  summary_jp(digits = 1, var1 = "Región", var2 = "Sexo", ft = TRUE, lan = "es")
```

## Regression plots

```{r}
#| id: plot-1
mod1 |>
  gg_jpoint()
```

Split by sex
```{r}
#| id: plot-2
mod1 |>
  gg_jpoint(facets = TRUE)
```

Split by region

```{r}
#| id: plot-3
mod3 |>
  gg_jpoint(facets = TRUE)
```

Hide data points
```{r}
#| id: plot-4
mod3 |>
  gg_jpoint(facets = TRUE, obs = FALSE)
```

Hide joinpoint(s) marker(s)
```{r}
#| id: plot-5
mod3 |>
  gg_jpoint(facets = TRUE, jp = FALSE)
```

Change colors and legend position using `ggplot2` functions
```{r}
#| id: plot-6
mod3 %>%
  gg_jpoint(facets = TRUE, jp = FALSE) +
  scale_color_viridis_d(name = "Sex:Region") +
  theme(legend.position = "bottom")
```

Change facet layout
```{r}
#| id: plot-7
mod3 %>%
  gg_jpoint(facets = TRUE) +
  facet_wrap(~group, ncol = 5) +
  scale_color_viridis_d(name = "Sex:Region") +
  theme(legend.position = "bottom")
```