---
title: "Introduction to joinpointR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| id: setup
#| include: false
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  dev = "ragg_png",
  comment = "#>"
)
```

Joinpoint regression is commonly used in epidemiology to identify changes in temporal trends. The joinpointR package provides tools to fit joinpoint regression models using the grid-search method, estimate summary statistics, and present results.

## Joinpoint regression models by group
The function `model_jp_grid()` allows to fit joinpoint regression models with a log-response by levels of up to two categorical variables, using the grid-search method and model selection based in the Bayesian Information Criteria (BIC).
For the examples, we are going to use a dataset with the HIV rates by sex in Argentina by jurisdiction between 2010 and 2022:

```{r}
#| message: false
#| id: example-data
# Load required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(joinpointR)

# Load example data
data(hiv_data)
```

### Model selection by one grouping variable

```{r}
#| id: fit-1
## Create a reduced dataset
data_sex <- hiv_data |>
  filter(admin == "ARG")

## Fit the joinpoint model
mods_sex <- model_jp_grid(
  data = data_sex,
  rate = hiv_rate,
  time = year,
  group = "sex"
)
```

### Model selection by two grouping variables
```{r}
#| id: fit-2
## Create a reduced dataset
data_admin <- hiv_data |>
  filter(between(admin, "Buenos Aires", "Chaco"))

## Fit the joinpoint model
mods_admin <- model_jp_grid(
  data = data_admin,
  rate = hiv_rate,
  time = year,
  group = c("admin", "sex")
)
```

### Change model selection method
By default `model_jp_grid()` uses the Bayesian Information Criteria (BIC) to select the best-fit model. Model selection method can be changed using the argument `method`.

```{r}
#| id: fit-3
# Fit the model using the weighted BIC
mods_wbic <- model_jp_grid(
  data = data_admin,
  rate = hiv_rate,
  time = year,
  group = c("sex", "admin"),
  method = "wbic"
)
```

## Summarise model results
### Bayesian Information Criteria
Regardless of the method selected, all the BIC metrics can be accesed using the function `bic_jp()`:

```{r}
#| id: bic
# BIC table for all the models
bic_jp(mods_admin)

# BIC table for a single model
bic_jp(mods_admin[1])
```

### Summary tables
The functions `get_summary()` provides a summary table containing:
- `model`: Model name.
- `jp`: The number of fitted joinpoints.
- `period`: The time segments defined by the observed joinpoints. Defaults to the whole period when no joinpoints were detected.
- `apc`,`apc_lower`, `apc_upper`, `apc_sig`: The Annual Percent Change (APC) with its confidence interval and significance stars.
- `aapc`,`aapc_lower`, `aapc_upper`, `apc_sig`: The Average Annual Percent Change (APC) with its confidence interval and significance stars. 

 ```{r}
 #| id: get_summary-1
 get_summary(mods_sex)
 ```

By default, `get_summary()` displays both the APC and AAPC, to show only the APC use:

```{r}
#| id: get_summary-2
get_summary(mods_sex, stats = "apc")

get_apc(mods_sex)
```

To display only the AAPC use:
```{r}
#| id: get_summary-3
get_summary(mods_sex, stats = "aapc")

get_aapc(mods_sex)
```

The confidence interval or significance stars can be hided using the argument `hide`:

```{r}
#|id: get_summary-4
# Hide the confidence interval
get_summary(mods_sex, hide = "ci")

# Hide the significance stars
get_summary(mods_sex, hide = "sig")
```

The confidence level defaults to 95% and can be changed using the argument `level.ci`:
```{r}
#| id: get_summary-5
get_summary(mods_sex, level.ci = .9)
```

Results can also be displayed as a `flextable`:

```{r}
#| id: get_summary-6
get_summary(mods_sex, as.ft = TRUE)
```

## Regression plots
### Default plot
One grouping variable
```{r}
#| id: ggjpoint-1
gg_jpoint(mods_sex)
```

Two grouping variables
```{r}
#| id: ggjpoint-2
gg_jpoint(mods_admin)
```

Change the facets layout
```{r}
#| id: ggjpoint-3
gg_jpoint(mods_admin, facets = "grid")
```

Reverse the facet grid
```{r}
#| id: ggjpoint-4
gg_jpoint(mods_admin, facets = "grid2")
```

Change the colors by APC trend

```{r}
#| id: ggjpoint-5
gg_jpoint(mods_admin, facets = "grid2", color.by = "trend")
```

Change the colors by time segment

```{r}
#| id: ggjpoint-6
gg_jpoint(mods_admin, facets = "grid2", color.by = "segment")
```

Change the colors by time period

```{r}
#| id: ggjpoint-7
gg_jpoint(mods_admin, facets = "grid2", color.by = "period")
```

Show Average Annual Percent Change (AAPC)
```{r}
#| id: ggjpoint-8
gg_jpoint(mods_admin, facets = "grid2", aapc = TRUE)
```

Hide joinpoint positions
```{r}
#| id: ggjpoint-9
gg_jpoint(mods_admin, facets = "grid2", jp = FALSE)
```


### Plot fitted lines
The observed log-rates can be hidden using `geom = "line"`
```{r}
#| id: ggjpoint-10
gg_jpoint(mods_admin, facets = "grid2", geom = "line")
```

Alternatively, the same results can be achieved using `gg_jpoint_line()`
```{r}
#| id: ggjpoint-11
gg_jpoint_line(mods_admin, facets = "grid2")
```

### Plot fitted areas
```{r}
#| id: ggjpoint-12
gg_jpoint(mods_sex, facets = "grid2", geom = "area")
```

The shortcut `gg_jpoint_area()`produces the same result
```{r}
#| id: ggjpoint-13
gg_jpoint_area(mods_sex, facets = "grid2")
```

### Change the default colors
By default, `gg_jpoint()` uses the default colorblind-friendly palette `"viridis"` for coloring the plots. Colorblind-friendly palettes can be changed using the functions `scale_cbpal_color()` and `scale_cbpal_fill()`.

```{r}
#| id: ggjpoint-14
gg_jpoint_line(mods_sex) +
  scale_cbpal_color(palette = "algae")

gg_jpoint_area(mods_sex, color.by = "trend") +
  scale_cbpal_fill(palette = "blue_fluoride")
```

The available colorblind-friendly palettes can be checked using `plot_cbpal()`:

```{r}
#| id: cbpal
plot_cbpal(type = "div")

plot_cbpal(series = "scico")
```